---
layout: post
title:  "Anode Pump Testing"
author: "Patrick M Kiel"
date: '2023-08-14'
categories: [research]
description: "Testing of the anode pump with a UV-Vis spectrophotometer to improve the lab-based mineral accretion technology setup."
output:
  md_document:
    variant: gfm
    preserve_yaml: TRUE
knit: (function(inputFile, encoding) {
  rmarkdown::render(inputFile, 
  encoding = encoding, 
  output_file = paste0("2023-08-14", "-",
                        gsub(pattern = "\\.Rmd$",
                              "", basename(inputFile))
                        ,".md"), 
  output_dir = "../_posts") })
always_allow_html: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)

#load packages
library(tidyverse)
library(kableExtra)
library(readxl)
library(ggpubr)
library(rstatix)

base_dir <- "C:/Users/pkiel/Github/notebook/" # i.e. where the jekyll blog is on the hard drive.
base_url <- "/notebook/" # subdirectory of patrickmkiel.com
fig_path <- "images/anodePump/" #add a new folder if you want to group output into a single foder

# Set base directories
knitr::opts_knit$set(base.dir = base_dir, base.url = base_url)

# Set figure directories
knitr::opts_chunk$set(fig.path = fig_path,
                    cache.path = '../cache/',
                    fig.align = 'center', out.width = '90%',
                    dpi=300,
                    message=FALSE, warning=FALSE,
                    cache = FALSE, results = "asis")
```

<!-- ```{js} -->
<!-- window.onload = function() { -->
<!--     //query string in the password -->
<!--     const urlParams = new URLSearchParams(window.location.search); -->
<!--     const pass = urlParams.get('pass') -->
<!--     document.getElementById("password").value = pass; -->
<!-- }; -->

<!-- function verify() { -->
<!-- set the password here -->
<!--   if (document.getElementById('password').value === 'password') { -->
<!--     document.getElementById('HIDDENDIV').classList.remove("hidden");  -->
<!--     document.getElementById('credentials').classList.add("hidden"); // Hide the div containing the credentials -->
<!--   } else { -->
<!--     alert('Invalid Password! You cannot view this content.'); -->
<!--     password.setSelectionRange(0, password.value.length); -->
<!--   } -->
<!--   return false; -->
<!-- } -->
<!-- ``` -->

```{css}
/*Change content Display */
<!-- .hidden { -->
<!--   display: none; -->
<!-- } -->

img {
margin: 0 auto;
}

table {
    width: 90%;
    border: 0px solid #fff;
    border-collapse: collapse;
    overflow-x: auto;
    margin: 0 auto;
    display: block;
}
```

```{r FigCaptions, echo=FALSE}

#Figure and Table Caption Numbering
capTabNo = 1; capFigNo = 1;

#Function to add the Table Number
capTab = function(x){
    x = paste0("Table ",capTabNo,". ",x)
    capTabNo <<- capTabNo + 1
    cat(x)
}

#Function to add the Figure Number
capFig = function(x){
    x = paste0("<h5>Figure ",capFigNo,". ",x,"</h5>")
    capFigNo <<- capFigNo + 1
    cat(x)
}
```

<!-- The password box -->
<!-- <div id="credentials"> -->
<!--   <input type="text" id="password" onkeydown="if (event.keyCode == 13) verify()" /> -->
<!--   <br/> -->
<!--   <input id="button" type="button" value="Show Content" onclick="verify()" /> -->
<!-- </div> -->

<!-- The content we want to show after password -->
<div id="HIDDENDIV" class="hidden" markdown="1">

<!-- Place all chunks, text, etc here as you would a normal RMarkdown document -->

# Overview

As part of my design to implement MAT in the laboratory, I am encapsulating the anode in a peristaltic pump to evacuate the anodic reactants, particularly chlorine gas. The goal of this anode pump testing is to calculate the efficiency rates of the pump and estimate how much chlorine gas may enter the system. 

As a reminder, here is the system:

<div class="sketchfab-embed-wrapper"> <iframe title="MATsetup" frameborder="0" allowfullscreen mozallowfullscreen="true" webkitallowfullscreen="true" allow="autoplay; fullscreen; xr-spatial-tracking" xr-spatial-tracking execution-while-out-of-viewport execution-while-not-rendered web-share src="https://sketchfab.com/models/1a9983e7bd37458b8810e8aee72231d1/embed" width="60%" style="height:35vh;"> </iframe> <p style="font-size: 13px; font-weight: normal; margin: 5px; color: #4A4A4A;"> <a href="https://sketchfab.com/3d-models/matsetup-1a9983e7bd37458b8810e8aee72231d1?utm_medium=embed&utm_campaign=share-popup&utm_content=1a9983e7bd37458b8810e8aee72231d1" target="_blank" rel="nofollow" style="font-weight: bold; color: #1CAAD9;"> MATsetup </a> by <a href="https://sketchfab.com/pkiel?utm_medium=embed&utm_campaign=share-popup&utm_content=1a9983e7bd37458b8810e8aee72231d1" target="_blank" rel="nofollow" style="font-weight: bold; color: #1CAAD9;"> pkiel </a> on <a href="https://sketchfab.com?utm_medium=embed&utm_campaign=share-popup&utm_content=1a9983e7bd37458b8810e8aee72231d1" target="_blank" rel="nofollow" style="font-weight: bold; color: #1CAAD9;">Sketchfab</a></p></div>

## Electrochemistry Half Reactions

1. @ the steel cathode, oxygen/water reduction
 * $O_{2(g)} + 2H_20_{(aq)} + 4e^- \rightarrow 4OH^-_{(aq)}$ (oxygen reduction)
 * $2H_2O_{(aq)} + 2e^- \rightarrow H_{2(g)} + 2(OH)^- _{(aq)}$ (water reduction)
 * $OH^-_{(aq)} + HCO_3^{2-} \iff CO_3^{2-} + H_2O_{(aq)}$
2. @ the platinum anode, oxidation of chloride
 * $2Cl^-_{(aq)} \rightarrow Cl_{2(g)} + 2e^-$
 
For steel, the potential needs to be < -0.356V/SSC (potential relative to silver-silver chloride electrode) to be below the corrosion domain. From -0.406 to -1.006V/SSC, the cathode is in the oxygen reduction domain; above 1.006V/SSC, the steel is in the water reduction domain. On a steel cathode, potentials more cathodic than −1.044 V/SCC, the pH value at the steel–seawater interface will exceed 9.0, which may favor the formation of brucite Mg(OH)2, which requires a minimum value of 9.3. 

Further, at these higher potentials, the rate of chloride oxidation increases. Thus, there is a two-fold advantage of operating at lower cathodic potentials between -0.4 and -1.0V/SSC to minimize chloride evolution and maintain within the oxygen reduction domain to precipiate calcium carbonate and not brucite. 

# Dye Testing

The end-goal of this test is to generate a table that looks like the following:

```{r, example table}
capTab("Idealized results for conceptual purposes only, not actual data")
tibble(sample = c("aquarium init", "effluent", "tank final"),
           `total volume` = c(75000, 2500, 72530),
           `dye volume` = c(0, 29.9, 0.1),
           `% evacuated` = c(NA, 99.7, 0.3)) %>%
  kbl()

```
This table shows that the initial volume of the aquarium is 75L and has no dye. 30mL of dye were added to the aquarium while the pump was running at 305mL/min for 8 minutes and 12 seconds, which decanted 2.5L into a separate container. Within that 2.5L, 29.9mL of the dye was collected and removed from the aquarium. The aquarium now has 72.53L of solution, which contains only 0.1mL of the dye that was not collected by the pump. Thus this idealized situation has a pump collection rat eof 99.7%

## First Test
The first test I ran on a brushed motor DC pump that was pumping at a rate of 30mL/min. This matched the 30mL/min rate of dye I planned to inject into the system.

<iframe width="560" height="315" src="https://www.youtube.com/embed/BzIQTHvdd6Q" title="YouTube video player" frameborder="0" allow="accelerometer; autoplay; clipboard-write; encrypted-media; gyroscope; picture-in-picture; web-share" allowfullscreen></iframe>

Immediately, I noticed problems that prevented further analysis. The dye was much denser than seawater so the majority of the dye was falling to the bottom of the tank. The chlorine gas will be buoyant bubbles that should follow the flow of water it is in, and so the dye is not a good proxy to test the effiency of the pump. Further, the pump rate was far too low to be of any use.

## Second Test

I used one of the new ERL2 peristaltic pumps which was getting a maximum rate of 305 $\pm$ 3 mL/min (mean $\pm$ 1SD of 5 tests). I then took the stock dye solution and measured its density to be approximately 1.16 $g \text{ } cm^-3$. I created 100mL of diluted dye stock to reach a density of 1.03 $g \text{ } cm^-3$, by diluting with RO water to match the density of the seawater in the aquarium. This density-mathced, diluted dye become the new dye standard from which I measured pump capture efficiency. I ran the following test twice, named round 1 and round 2.

<iframe width="560" height="315" src="https://www.youtube.com/embed/gZEQghycPDY" title="YouTube video player" frameborder="0" allow="accelerometer; autoplay; clipboard-write; encrypted-media; gyroscope; picture-in-picture; web-share" allowfullscreen></iframe>

First, I created a calibration curve, to linearly model the Beer-Lambert Law, with the formula $A=\epsilon lC$, where A is absorbance, $\epsilon$ is the molar absorption coefficient, l is the path length of the cuvette (1 cm), and C is concentration of the sample/standard. Concentration was expressed as volume of dye per volume solution.

Since the molar absorption coefficient and path length is constant, the terms $\epsilon * l$ can be combined into the constant m, to produce the simplified equation A=mC. Standard concentrations bracketting the expected concentration of the samples were prepared and plotted against their measured absorbances to calculate the slope m.

This is done simply in R using the lm function to generate a linear model and the geom_smooth funciton of ggplot with the method="lm" to plot the data.

For the standards, I expected the best case scenario to be 29mL dye/2500 mL solution = 1.2% and 1mL dye/72530 mL to be 0.001%, likely below the detectable limit. At the worst case, no dye would be collected and the concentration of the aquarium after the pump removed exclusively 2.5L of seawater would be 30mL dye/72530mL solution = 0.04%. So I created standard solutions in triplicate from 0.03 - 3.3 % to cover all scenarios.
```{r eval=F, echo=T, include=T}
calibration <- calibration %>%
  mutate(concentration = dye/(dye+RO))

calibration_lm <- lm(abs~concentration, data = calibration)

calibration %>%
  gplot(aes(concentration,abs)) +
  geom_point() +
  geom_smooth(se=F,method="lm")
```

```{r  class.source = 'fold-show', calibration}
dat <- read.csv("~/Grad School/mineralAccretionTechnology/anodePumpTest071923/dyeTest072623.csv")

calibration <- dat %>%
  filter(round == "calibration") %>%
  mutate(concentration = dye/(dye+RO))

calibration_lm <- lm(abs~concentration,
                     data = calibration)

#summary(calibration_lm)

capFig("Dye standard curve absorbance at 480nm")
calibration %>%
  ggplot(aes(concentration, abs)) +
  geom_point() +
  geom_smooth(se=F, method="lm") +
  theme_bw() +
  labs(x = "Concentration (mL dye/mL sol)",
       y = "Absorbance",
       title = "A = 8.178 * C")

capFig("Precision increases as concentration increaes")
calibration %>% 
  group_by(concentration) %>% 
  summarise(mean = mean(abs), sd = sd(abs), cv= sd/mean) %>%
  ungroup() %>%
  ggplot(aes(concentration, cv)) +
  geom_point() +
  theme_bw() +
  labs(x = "Concentration (mL dye/mL sol)",
     y = "Coefficient of Variation")
```
There was high degrees of precision, with a mean coefficient of variation (CV) of 0.0504%. Nevertheless, precision decreased as concentration decreased, evidenced by the increase of coefficient of variation at low concentrations.

The linear function to determine concentration from absorbance fit the Beer-Lambert assumptions, including an intercept of approximately 0 and a high coefficient of determination ($R^2$) of 0.994.  

```{r, rd1}
rd1 <- dat %>%
  filter(round == "rd1") %>%
  group_by(sample_id) %>%
  summarise(abs = mean(abs), `sol volume`=mean(volume)) %>%
  ungroup()

rd1$sample_id <- factor(rd1$sample_id,
                               levels = c('dye', 'tankInit', 'tankDyeImmediate',
                                          'effluent', 'tank30min')) 

capTab("Round 1: Average absorbance from triplicate readings")
rd1 %>%
  arrange(factor(sample_id)) %>%
  kbl()

capTab("Round 1: Subtracted tank_init, calculate C and dye.vol")
rd1 %>%
  mutate(abs = ifelse(sample_id != "dye",
                      abs-0.002,
                      abs),
         concentration = abs*calibration_lm$coefficients[2],
         dye.vol = `sol volume`*concentration) %>%
  arrange(factor(sample_id)) %>%
  kbl()
```
These numbers don't make sense. While concentration is highest in effluent as expected, it's dye vol is estimated at 1.286L, which is well above the 30mL I injected. I'm wondering if this is a problem of units? I'm using dye/solution concentration %, but Beer-Lambert is meant for molar concentrations. I'm looking into this some more.

</div>
