---
layout: post
title:  "ULink 2022 Genotype x OA Growth Analysis"
author: "Patrick M Kiel"
date: '`r Sys.Date()`'
categories: [research]
description: "Growth analysis of corals in the University of Miami ULINK Genotype x ocean acidifcation experiment to test for mechanisms of resilience to global change within the critically endangered Acropora cervicornis."
output:
  md_document:
    variant: gfm
    preserve_yaml: TRUE
knit: (function(inputFile, encoding) {
  rmarkdown::render(inputFile, 
  encoding = encoding, 
  output_file = paste0(Sys.Date(), "-",
                        gsub(pattern = "\\.Rmd$",
                              "", basename(inputFile))
                        ,".md"), 
  output_dir = "../_posts") })
always_allow_html: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)

#load packages
library(tidyverse)
library(kableExtra)
library(googlesheets4)
library(seacarb)
library(emmeans)
library(ggpubr)
library(rstatix)

#selet the authorized tokens
googledrive::drive_auth("p.kiel98@gmail.com")
gs4_auth("p.kiel98@gmail.com")


base_dir <- "C:/Users/pkiel/Github/notebook/" # i.e. where the jekyll blog is on the hard drive.
base_url <- "/notebook/" # subdirectory of patrickmkiel.com
fig_path <- "images/ulinkGrowth2022/" #add a new folder if you want to group output into a single foder

# Set base directories
knitr::opts_knit$set(base.dir = base_dir, base.url = base_url)

# Set figure directories
knitr::opts_chunk$set(fig.path = fig_path,
                    cache.path = '../cache/',
                    fig.align = 'center', out.width = '60%',
                    message=FALSE, warning=FALSE,
                    cache = FALSE)
```

```{js}
window.onload = function() {
    //query string in the password
    const urlParams = new URLSearchParams(window.location.search);
    const pass = urlParams.get('pass')
    document.getElementById("password").value = pass;
};


function verify() {
  <!-- set the password here -->
  if (document.getElementById('password').value === 'ulink') {
    document.getElementById('HIDDENDIV').classList.remove("hidden"); 
    document.getElementById('credentials').classList.add("hidden"); // Hide the div containing the credentials
  } else {
    alert('Invalid Password! You cannot view this content.');
    password.setSelectionRange(0, password.value.length);
  }
  return false;
}
```

```{css}
/*Change content Display */
.hidden {
  display: none;
}
```

<!-- The password box -->
<div id="credentials">
  <input type="text" id="password" onkeydown="if (event.keyCode == 13) verify()" />
  <br/>
  <input id="button" type="button" value="Show Content" onclick="verify()" />
</div>

<!-- The content we want to show after password -->
<div id="HIDDENDIV" class="hidden" markdown="1">

<!-- Place all chunks, text, etc here as you would a normal RMarkdown document -->

# Overview
Here I review the growth we have observed in our experiment. The total growth was less than we anticipated, but we still produced enough new skeleton with significant differences in growth rates and sensitivities to proceed forward with most of our tests. 

# Calcification Analysis

```{r load buoyant weight data}
#load data from google drive
bw <- googledrive::drive_ls(path = "~/GradSchool/ULink_OAxGenotype/",
                            pattern = "buoyantWeight$") %>%
  read_sheet(col_types = "Dcdddc") %>%
  mutate(tag = as.character(tag)) %>%
  drop_na(mass)

#load tank assignment + genotype data
assignment <- read.csv('~/Grad School/ULink 2022/tankAssignment.csv') %>%
  filter(treatment != "") %>%
  mutate(tag = as.character(tag),
         tank = as.factor(tank))

bw <- bw %>%
  left_join(assignment, by="tag") %>%
  arrange(tank, genotype,tag,date) %>%
  filter(date > as.Date("2022-03-15"))
  

#calculate weight in air
bw <- bw %>%
  mutate(
    #calculate sw density (g/cm3) from YSI, p taken as std atm p
    density = rho(S = salinity,
                          T = temp)/1000) %>%
  mutate(
    #calculate weight in air using Jokiel et al. 1978 eqn
    mass_a = mass/(1-density/2.93)) %>%
  select(-c(mass,temp,salinity, density))

#calculate weekly growth rate for QA/QC
weeklyGrowth <- bw %>%
  #week of 05/27/2022 had lots of QC issues, removing data
  filter(!date %in% c(as.Date('2022-05-27'))) %>%
  drop_na(genotype, mass_a) %>%
  group_by(tag = as.numeric(tag)) %>%
  mutate(weekNum = as.numeric(date-first(date))/7,
            newGrowth.mg = ifelse(mass_a!=first(mass_a),
                                  (mass_a - lag(mass_a))*1000,
                                  NA),
            weeklyG.day = ifelse(mass_a!=first(mass_a),
                                  #average daily growth for the week
                                  #mg/g/day, normalized to initial mass
                                  newGrowth.mg /
                                    first(mass_a)/as.numeric(date-lag(date)),
                                  NA),
            weeklyG.week = ifelse(mass_a!=first(mass_a),
                                      #average weekly growth
                                      #mg/g/week, normalized to initial mass
                                      newGrowth.mg/
                                 first(mass_a)/(as.numeric(date-lag(date))/7),
                                      NA)) %>%
  #remove tags that have no weekly growth, i.e. died early on
  group_by(tag) %>%
  filter(!all(is.na(weeklyG.day)))

#corals to check
#weird growth
#10, 63, 103,
#likely broke
#120, 49

#calculate total growth
totalGrowth <- bw %>%
  #set starting date
  filter(date > as.Date("2022-04-15")) %>%
  #remove standards
  drop_na(genotype,mass_a) %>%
  group_by(tag,genotype,tank, treatment) %>%
  summarise(days = as.numeric(last(date)-first(date)),
            newGrowth.mg = (last(mass_a) - first(mass_a))*1000,
            initMass = first(mass_a),
            dailyG = #mg/g/day, normalized to initial mass
                           newGrowth.mg/initMass/days,
            .groups="drop")
```


```{r growth graphs}
weeklyGrowth %>%
  group_by(weekNum,genotype) %>%
  summarise(weeklyG = mean(weeklyG.week, na.rm = T),
            se = sd(weeklyG.week, na.rm = T)/sqrt(n())) %>%
  filter(!is.nan(weeklyG)) %>%
  ggplot(aes(weekNum,weeklyG)) +
  geom_line(aes(color=genotype)) +
  geom_ribbon(aes(ymin = weeklyG - se,
                    ymax = weeklyG + se,
                    fill=genotype),
                alpha=0.3) + 
  labs(title = "Avg Weekly Growth Per Week by Genotype ± se") +
  scale_x_continuous("Week",
                     breaks = c(seq(1,
                                    max(weeklyGrowth$weekNum, na.rm = T),
                                    by=1))) +
  scale_y_continuous("Avg Weekly Growth Rate (mg/g/week)",
                     limits = c(0,40),
                     breaks = c(seq(0,40,by=10)),
                     expand = c(0, 0)) +
  theme_classic()

weeklyGrowth %>%
  group_by(weekNum,treatment) %>%
  summarise(weeklyG = mean(weeklyG.week, na.rm = T),
            se = sd(weeklyG.week, na.rm = T)/sqrt(n())) %>%
  filter(!is.nan(weeklyG)) %>%
  ggplot(aes(as.integer(weekNum),weeklyG)) +
  geom_line(aes(color=treatment)) +
  geom_ribbon(aes(ymin = weeklyG - se,
                    ymax = weeklyG + se,
                    fill=treatment),
                alpha=0.3) + 
  labs(title = "Avg Weekly Growth Per Week by Treatment Group ± se") +
  scale_x_continuous("Week",
                     breaks = c(seq(1,
                                    max(weeklyGrowth$weekNum, na.rm = T),
                                    by=1))) +
  scale_y_continuous("Avg Weekly Growth Rate (mg/g/week)",
                     limits = c(0,40),
                     breaks = c(seq(0,40,by=10)),
                     expand = c(0, 0)) +
  theme_classic()
```

Following April 15 (Weeky 5), the declining health of the corals stabilized and began to split amongst treatment groups.

```{r calcification graphs}
totalGrowth %>%
  ggplot(aes(treatment, dailyG)) +
  geom_boxplot() +
  labs(x = "Treatment",
       y = "Avg Daily Growth Rate (mg/g/day)",
       title = "Avg Daily Growth byTreatment") +
  theme_classic()


totalGrowth %>%
  ggplot(aes(genotype, dailyG)) +
  geom_boxplot(aes(fill=treatment)) +
              #if quantile lines desired: draw_quantiles = c(0.25, 0.5, 0.75)) +
  labs(x = "Genotype",
       y = "Avg Daily Growth Rate (mg/g/day)",
       title = "Avg Daily Growth by Genotype and Treatment") +
  theme_classic()
```

There is some obvious genet-specific responses.

1. Cheetos-B calcification rate was nearly identical across HCO2 and LCO2 groups. This genet also had high initial mortality and the worst survivorship rate throughout the experiment. It is entirely possible that this genotype did not do well in the aquariums and its diminished calcification rate is an effect of overall health and not treatment.

2. SI-A and AC-2 had the highest average calcification rates and there was no significant difference between these two genotypes. However, when you look at the effect of treatment within these genotypes (sensitivity), there is significant differences between them.

3. When looking at just controls, the only significant different genotype is Cheetos-B. Thus, there is a difference in sensitivity to OA but no observable differences in ambient conditions. 

4. The relative rankings extracted from the ol' AcDC are SI-A ~ MB-C > AC-2 > Cheetos-B. The data collected here fits within that framework, yet reveals intraspecifc differences in sensitivities similar to Enochs et al. (2018). Genet identities are unknown for that paper though.

```{r calcification stats}

groupMeans <- totalGrowth %>%
  group_by(treatment) %>%
  get_summary_stats(dailyG, type = "mean_sd")

groupMeans %>%
  select(-variable) %>%
      kbl(align = 'c',
    digits = 4,
    caption = "Means of Treatment") %>%
  kable_classic() %>%
  row_spec(0, bold = T)

tTestResults <- totalGrowth %>% 
  t_test(dailyG ~ treatment, var.equal = T) %>%
  add_significance()

tTestResults %>%
  kbl(align = 'c',
    digits = 4,
    caption = "Stats of Treatment",
    escape=F) %>%
  kable_classic() %>%
  row_spec(0, bold = T)

cohensResults <- totalGrowth %>% 
  cohens_d(dailyG ~ treatment, var.equal = T) 

cohensResults %>%
      kbl(align = 'c',
    digits = 4,
    caption = "Effect Size of Treatment") %>%
  kable_classic() %>%
  row_spec(0, bold = T)

#fixed effects model
mod <- totalGrowth %>%
  #filter(genotype!='Cheetos-B') %>%
  aov(dailyG ~ treatment*genotype,
                 data=.) 

mod %>%
  TukeyHSD() %>%
  broom::tidy() %>%
    filter(contrast %in% c('LCO2-HCO2',
                        'Cheetos-B-AC-2',
                        'MB-C-AC-2,',
                        'SI-A-AC-2',
                        'MB-C-Cheetos-B',
                        'SI-A-Cheetos-B',
                        'SI-A-MB-C',
                        'LCO2:AC-2-HCO2:AC-2',
                        'LCO2:Cheetos-B-HCO2:Cheetos-B',
                        'LCO2:MB-C-HCO2:MB-C',
                        'LCO2:SI-A-HCO2:SI-A')) %>%
  select(term,contrast,adj.p.value) %>%
  mutate(significance = case_when(adj.p.value <0.001 ~ "xxx",
                                  adj.p.value <0.01 ~ "xx",
                                  adj.p.value <0.05 ~ "x",
                                  TRUE ~ "NS"),
         adj.p.value = round(adj.p.value, digits=4)) %>%
  arrange(adj.p.value) %>%
    kbl(align = 'c',
    digits = 4,
    caption = "TukeyHSD Results of Anova",
    escape = TRUE) %>%
  kable_classic() %>%
  row_spec(0, bold = T)
  
multcomp::cld(emmeans(mod,
                      list(pairwise ~ treatment*genotype),
                      adjust = "tukey"),
              Letters = letters) %>%
  select(treatment, genotype, significance = .group) %>%
    kbl(align = 'c',
    digits = 4,
    caption = "Significance Letter Groups") %>%
  kable_classic() %>%
  row_spec(0, bold = T)
```

The mean calcification rate in the HCO2 group was mean `r round(mean(totalGrowth$dailyG[which(totalGrowth$treatment=='HCO2')]),3)`  (SD = `r round(sd(totalGrowth$dailyG[which(totalGrowth$treatment=='HCO2')]),3)`) mg/g/day, whereas the mean in the LCO2 group was `r round(mean(totalGrowth$dailyG[which(totalGrowth$treatment=='LCO2')]),3)` (SD = `r round(sd(totalGrowth$dailyG[which(totalGrowth$treatment=='LCO2')]),3)`). A Student two-samples t-test showed that the difference was statistically significant, t(`r tTestResults$df`) = `r round(tTestResults$statistic,3)`, p < 0.0001, d = `r round(cohensResults$effsize,3)`. Thus, the ocean acidification group saw on average a `r round((groupMeans$mean[2]-groupMeans$mean[1])/groupMeans$mean[2]*100,0)`% reduction in calcification rates.

## Powder Available

```{r powder available}
totalGrowth %>%
  ggplot(aes(x=genotype, y=newGrowth.mg)) +
  geom_point(aes(color=treatment)) +
  scale_y_continuous(limits = c(0,1000)) +
  geom_hline(yintercept = 50, color="red") +
  geom_hline(yintercept = 120, color="orange") +
  geom_hline(yintercept = 500, color="green") +
  labs(title = "Amount of New CaCO3 Precipated",
       y = "Precipated Skeleton (mg)") +
  theme_classic()
```

The amount of new aragonite precipated is visualized above. Horizontal lines denote thresholds for tests: >500mg = green (complete suite including XRD), >120 mg = orange (complete suite sans XRD), >50mg = red (TGA and isotope only).

# Linear Extension

```{r import LE data}
height <- read.csv('~/Grad School/ULink 2022/initialHeights.csv') %>%
  mutate(date=as.Date("2022-02-25")) %>%
  select(tag,height=height.mm,date) %>%
  left_join(read.csv('~/Grad School/ULink 2022/height06102022.csv') %>%
              mutate(date=as.Date("2022-06-10"),
                     .after="height"),
            by="tag") %>%
  select(tag:date.y) %>%
  left_join(read.csv('~/Grad School/ULink 2022/height07152022.csv'),
            by="tag") %>%
  mutate(height.z=height,
         date.z=as.Date("2022-07-15")) %>%
  select(-height) %>%
  pivot_longer(!tag,
               names_to = '.value',
               names_pattern = '(.*)\\.',
               values_drop_na = T) %>%
  drop_na() %>%
  arrange(tag,date) %>%
  group_by(tag) %>%
  summarise(init.mm = min(height),
            fin.mm = max(height),
            days = as.numeric(date[which.max(height)]-first(date))) %>%
  filter(days!=0) %>%
  mutate(prod = (fin.mm-init.mm)/(init.mm/10)/days,
         LE = fin.mm-init.mm) %>%
  left_join(assignment %>%
              mutate(tag = as.numeric(tag)),
            by="tag")

```

```{r LE graphs}
height %>%
  ggplot(aes(treatment, prod)) +
  geom_boxplot() +
  labs(x = "Treatment",
       y = "Avg Daily LE Rate (mm/cm/day)",
       title = "Avg Daily Linear Extension by Treatment") +
  theme_classic()


height %>%
  ggplot(aes(genotype, prod)) +
  geom_boxplot(aes(fill=treatment)) +
              #if quantile lines desired: draw_quantiles = c(0.25, 0.5, 0.75)) +
  labs(x = "Genotype",
       y = "Avg Daily LE Rate (mm/cm/day)",
       title = "Avg Daily Linear Extension by Genotype and Treatment") +
  theme_classic()
```

```{r LE stats}

groupMeans <- height %>%
  group_by(treatment) %>%
  get_summary_stats(prod, type = "mean_sd")

groupMeans %>%
  select(-variable) %>%
      kbl(align = 'c',
    digits = 4,
    caption = "Means of Treatment") %>%
  kable_classic() %>%
  row_spec(0, bold = T)

tTestResults <- height %>% 
  t_test(prod ~ treatment, var.equal = T) %>%
  add_significance()

tTestResults %>%
  kbl(align = 'c',
    digits = 4,
    caption = "Stats of Treatment",
    escape=F) %>%
  kable_classic() %>%
  row_spec(0, bold = T)

cohensResults <- height %>% 
  cohens_d(prod ~ treatment, var.equal = T) 

cohensResults %>%
      kbl(align = 'c',
    digits = 4,
    caption = "Effect Size of Treatment") %>%
  kable_classic() %>%
  row_spec(0, bold = T)

#fixed effects model
mod <- height %>%
  aov(prod ~ treatment*genotype,
                 data=.) 

mod %>%
  TukeyHSD() %>%
  broom::tidy() %>%
  filter(contrast %in% c('LCO2-HCO2',
                        'Cheetos-B-AC-2',
                        'MB-C-AC-2,',
                        'SI-A-AC-2',
                        'MB-C-Cheetos-B',
                        'SI-A-Cheetos-B',
                        'SI-A-MB-C',
                        'LCO2:AC-2-HCO2:AC-2',
                        'LCO2:Cheetos-B-HCO2:Cheetos-B',
                        'LCO2:MB-C-HCO2:MB-C',
                        'LCO2:SI-A-HCO2:SI-A')) %>%
  select(term,contrast,adj.p.value) %>%
  mutate(significance = case_when(adj.p.value <0.001 ~ "xxx",
                                  adj.p.value <0.01 ~ "xx",
                                  adj.p.value <0.05 ~ "x",
                                  TRUE ~ "NS"),
         adj.p.value = round(adj.p.value, digits=4)) %>%
  arrange(adj.p.value) %>%
    kbl(align = 'c',
    digits = 4,
    caption = "TukeyHSD Results of Anova",
    escape = F) %>%
  kable_classic() %>%
  row_spec(0, bold = T)
  
multcomp::cld(emmeans(mod,
                      list(pairwise ~ treatment*genotype),
                      adjust = "tukey"),
              Letters = letters) %>%
  select(treatment, genotype, significance = .group) %>%
    kbl(align = 'c',
    digits = 4,
    caption = "Significance Letter Groups") %>%
  kable_classic() %>%
  row_spec(0, bold = T)
```

The mean linear extension rate in the HCO2 group was mean `r round(mean(height$prod[which(height$treatment=='HCO2')]),3)`  (SD = `r round(sd(height$prod[which(height$treatment=='HCO2')]),3)`) mm/cm/day, whereas the mean in the LCO2 group was `r round(mean(height$prod[which(height$treatment=='LCO2')]),3)` (SD = `r round(sd(height$prod[which(height$treatment=='LCO2')]),3)`). A Student two-samples t-test showed that the difference was not statistically significant, t(`r tTestResults$df`) = `r round(tTestResults$statistic,3)`, p =0.124, d = `r round(cohensResults$effsize,3)`. 

```{r total extension}
height %>%
  ggplot(aes(x=genotype, y=LE)) +
  geom_point(aes(color=treatment)) +
  scale_y_continuous(limits = c(0,30)) +
  geom_hline(yintercept = mean(height$LE),
             color="red") +
  labs(title = "Amount of New Skeleton and Mean New LE (Red Line)",
       y = "Linear Extension (mm)") +
  theme_classic()
```

# Takeaways and Next Steps

Overall growth was less than hoped for. However, there is enough new skeleton for nearly all the powder tests that we want to conduct: FTIR, TGA, boron isotopes, and Raman. There is enough powder to run XRD on some samples from genotypes AC-2 (OA=2, control=8) and SI-A (control=5) and a single sample of Cheetos-B. The lack of even replication though may preclude analysis of XRD. For nanoindentation tests, we should have enough new growth to run on a majority of samples.

Linear extension was measured by maximum vertical height as measured with calipers. We additionally have initial 3d scans of all corals and post 3d scans of a subset of 48 corals (n=3 per genotype per tank). From this data, we can extract surface area to volume ratios and see how this changed among genotypes and treatments. This analysis still needs to be done. We can also more accurately measure total linear extension of the corals which may have curving or branching morphologies.

## Plating of Tissue
Among control corals, I visually noticed significant plating of tissue and a veneer of aragonite above the acrylic tags. This was almost completely absent in OA corals. I can take photos of each coral and calculate surface area of plating.

## CT Scanning
The micro-ct scan is currently out of service. We can use the large ct-scanner to determine bulk density. The scanner has a resolution of 0.1mm/scan so we can measure the density of the new growth. This growth is isolated to the highly variable apical tips which may cause some problems. <a href="https://patrickmkiel.com/notebook/research/LangdonSkeletonAnalysis/?pass=acidification" target="_blank">See this post which discusses the ct-scanning analysis of apical tips done on Langdon's OA corals.</a> 


</div>
