---
layout: post
title:  "ULink 2022 Genotype x OA Growth Analysis"
author: "Patrick M Kiel"
date: '2022-08-03'
categories: [research]
description: "Growth analysis of corals in the University of Miami ULINK Genotype x ocean acidifcation experiment to test for mechanisms of resilience to global change within the critically endangered Acropora cervicornis."
output:
  md_document:
    variant: gfm
    preserve_yaml: TRUE
knit: (function(inputFile, encoding) {
  rmarkdown::render(inputFile, 
  encoding = encoding, 
  output_file = paste0("2022-08-03-",
                        gsub(pattern = "\\.Rmd$",
                              "", basename(inputFile))
                        ,".md"), 
  output_dir = "../_posts") })
always_allow_html: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)

#load packages
library(tidyverse)
library(kableExtra)
library(googlesheets4)
library(seacarb)
library(emmeans)
library(ggpubr)
library(rstatix)
library(openxlsx)
library(lubridate)
library(scales)
library(wesanderson)

base_dir <- "C:/Users/pkiel/Github/notebook/" # i.e. where the jekyll blog is on the hard drive.
base_url <- "/notebook/" # subdirectory of patrickmkiel.com
fig_path <- "images/ulinkGrowth2022/" #add a new folder if you want to group output into a single foder

# Set base directories
knitr::opts_knit$set(base.dir = base_dir, base.url = base_url)

# Set figure directories
knitr::opts_chunk$set(fig.path = fig_path,
                    cache.path = '../cache/',
                    fig.align = 'center', out.width = '90%',
                    message=FALSE, warning=FALSE,
                    cache = FALSE, results = 'asis')

#set default color paletes
options(ggplot2.discrete.fill=wes_palette("Darjeeling1"))
options(ggplot2.discrete.colour=wes_palette("Darjeeling1"))
```

```{js}
window.onload = function() {
    //query string in the password
    const urlParams = new URLSearchParams(window.location.search);
    const pass = urlParams.get('pass')
    document.getElementById("password").value = pass;
};


function verify() {
  <!-- set the password here -->
  if (document.getElementById('password').value === 'ulink') {
    document.getElementById('HIDDENDIV').classList.remove("hidden"); 
    document.getElementById('credentials').classList.add("hidden"); // Hide the div containing the credentials
  } else {
    alert('Invalid Password! You cannot view this content.');
    password.setSelectionRange(0, password.value.length);
  }
  return false;
}
```

```{css}
/*Change content Display */
.hidden {
  display: none;
}
```


```{r FigCaptions, echo=FALSE}

#Figure and Table Caption Numbering
capTabNo = 1; capFigNo = 1;

#Function to add the Table Number
capTab = function(x){
    x = paste0("Table ",capTabNo,". ",x)
    capTabNo <<- capTabNo + 1
    cat(x)
}

#Function to add the Figure Number
capFig = function(x){
    x = paste0("<h5>Figure ",capFigNo,". ",x,"</h5>")
    capFigNo <<- capFigNo + 1
    cat(x)
}
```


<!-- The password box -->
<div id="credentials">
  <input type="text" id="password" onkeydown="if (event.keyCode == 13) verify()" />
  <br/>
  <input id="button" type="button" value="Show Content" onclick="verify()" />
</div>

<!-- The content we want to show after password -->
<div id="HIDDENDIV" class="hidden" markdown="1">

<!-- Place all chunks, text, etc here as you would a normal RMarkdown document -->

# Overview
Here I review the growth we have observed in our experiment. The total growth was less than we anticipated, but we still produced enough new skeleton with significant differences in growth rates and sensitivities to proceed forward with most of our tests.

# Treatment Conditions

## Labview Data

```{r load labview}
#load data
pH <- list.files(pattern = "(.*)(_PHData.xlsx$)",
                path = "~/Grad School/ULink 2022/ERLLabView/",
                full.names = T) %>%
  #remove files that begin with ~
  magrittr::extract(which(str_detect(.,"~",negate = T))) %>%
  lapply(., function(x) {
    x %>%
      readxl::read_excel(.,
                         col_types = c("text", "numeric", "numeric", 
                                       "numeric", "numeric", "numeric", 
                                       "numeric", "numeric", "numeric", 
                                       "numeric", "numeric", "numeric", 
                                       "numeric", "numeric", "numeric", 
                                       "numeric", "numeric"))
  }) %>%
  bind_rows() %>%
  distinct() %>%
  mutate(time = convertToDateTime(Time, origin = "1904-01-01",
                                  tz = "America/New_York")) %>%
  select(time, 
         `13` = `Untitled 12`,
         `14` = `Untitled 13`,
         `15` = `Untitled 14`,
         `16` = `Untitled 15`) %>%
  pivot_longer(cols = `13`:`16`,
               names_to = "tank",
               values_to = "pH")

temp <- list.files(pattern = "_TemperatureData.xlsx$",
                   path = "~/Grad School/ULink 2022/ERLLabView/",
                   full.names = T) %>%
  #remove files that begin with ~
  magrittr::extract(which(str_detect(.,"~",negate = T))) %>%
  lapply(., function(x) {
    x %>%
      readxl::read_excel(.,
                         col_types = c("text", "numeric", "numeric", 
                                       "numeric", "numeric", "numeric", 
                                       "numeric", "numeric", "numeric", 
                                       "numeric", "numeric", "numeric", 
                                       "numeric", "numeric", "numeric", 
                                       "numeric", "numeric"))
  }) %>%
  bind_rows() %>%
  distinct() %>%
  mutate(time = convertToDateTime(Time, origin = "1904-01-01",
                                  tz = "America/New_York")) %>%
  select(time, 
         `13` = `Untitled 12`,
         `14` = `Untitled 13`,
         `15` = `Untitled 14`,
         `16` = `Untitled 15`) %>%
  pivot_longer(cols = `13`:`16`,
               names_to = "tank",
               values_to = "temp")

setPoints <- readxl::read_excel('~/Grad School/ULink 2022/pHFinalOscillation.xlsx') %>%
  pivot_longer(control:treat,
               names_to = "type",
               values_to = "pH.set")

labViewData <- full_join(pH,temp, by=c("time", "tank")) %>%
  arrange(time) %>%
  #convert UTC to EST
  mutate(time = with_tz(time, 
                        "America/New_York"),
         hour = hour(time),
         type = case_when(
           tank %in% c(13,15) ~ "treat",
           TRUE ~ "control")) %>%
  #add in set points
  left_join(setPoints,
            by=c("type","hour"="time")) %>%
  select(-c(type,hour))
```

```{r diel plot}
capFig("10-minute averaged ERL Log")

labViewData %>%
  #only use data once treatment conditions met 4/1
  filter(date(time) > as.Date("2022-04-01")) %>%
  drop_na() %>%
  mutate(hms = as.POSIXct(format(time,
                           format="%H:%M:%S"), 
                           format = "%H:%M:%S",
                           tz = "America/New_York"),
          #scale temp to fit on same graph
          temp = temp *  0.0625 + 6.45) %>%
  group_by(hms = round_date(hms, "10 minutes"), tank) %>%
  summarise(pH.mean = mean(pH,na.rm=T),
            pH.sd = sd(pH,na.rm=T),
            temp.mean = mean(temp,na.rm=T),
            temp.sd = sd(temp,na.rm=T)) %>%
  ggplot(aes(hms)) +
  geom_smooth(aes(y=pH.mean, color="pH"),
              se = F, span=0.5) +
  geom_ribbon(aes(ymin=pH.mean-pH.sd,
                  ymax=pH.mean+pH.sd,
                  fill="pH"), alpha=0.3) + 
  geom_line(aes(y=temp.mean, color="Temp")) +
  geom_ribbon(aes(ymin=temp.mean-temp.sd,
                ymax=temp.mean+temp.sd,
                fill="temp"),alpha=0.3) + 
  scale_y_continuous(
    # Features of the first axis
    name = "pH",
    breaks = seq(7.8,8.2,by=0.2),
    minor_breaks = seq(7.7,8.2,by=0.05),
    limits = c(7.7,8.2),
    # Add a second axis and specify its features
    sec.axis = sec_axis(~(.- 6.45) / 0.0625, 
                        name="Temperature")
  ) + 
  scale_x_datetime("Hour",
                   limits = as.POSIXct(strptime(c("00:00",
                                                  "23:59"),
                                                format = "%H:%M")),
                   breaks = seq(as.POSIXct("00:00", format="%H:%M"),
                                as.POSIXct("24:00", format="%H:%M"),
                                by = "3 hours"),
                   #expand = c(0,0),
                   labels = date_format("%H")) +
  facet_wrap(~tank) +
  theme_bw() +
  guides(fill = FALSE)
```

The peaks in the standard deviation are almost certainly caused by aquarium cleaning days when corals are removed into a separate bath and the sensors are capped causing logging errors. CO2 injection is turned off during these times so the aquariums themselves are not experiencing the conditions that the logged data are suggesting. The following graph filters out these spiked values which were programmatically identified by occurring during scheduled cleaning times and affecting multiple aquariums at once since cleaning occurred at the same time for all aquariums.

```{r cleaned diel plot}
#gather times affected by cleaning of aquarium
#corals are placed in a separate bath and are not affected by the rapid swing
#cleaned during the week, between 9a and 6p
#defined when multiple tanks affected
cleaningTimes <- labViewData %>%
  filter(date(time) > as.Date("2022-04-01")) %>%
  group_by(tank) %>%
  filter(weekdays(time) %in% weekdays(ISOdate(1, 1, 1:5)) &
           between(hour(time),9,18) &
           abs(pH-pH.set)>0.05) %>%
  ungroup() %>%
  group_by(time = round_date(time, "30 minutes")) %>%
  filter(length(unique(tank))>1)

capFig("Cleaned 10-minute Averaged ERL Log")
labViewData %>%
  #only use data once treatment conditions met 4/1
  filter(date(time) > as.Date("2022-04-01")) %>%
  drop_na() %>%
  filter(!round_date(time, "30 minutes") %in% cleaningTimes$time) %>%
  #durafet swapped T14 4/22 9:30a-16:30
  filter(!(between(time,
                 as.POSIXct("2022-04-22 09:30:00",
                            tz="America/New_York"),
                 as.POSIXct("2022-04-22 16:30:00",
                            tz="America/New_York")) & tank==14)) %>%
  #durafet swapped T15 7/8 
    filter(!(between(time,
                 as.POSIXct("2022-07-08 14:13:00",
                            tz="America/New_York"),
                 as.POSIXct("2022-07-08 19:30:00",
                            tz="America/New_York")) & tank==15)) %>%
  mutate(hms = as.POSIXct(format(time,
                           format="%H:%M:%S"), 
                           format = "%H:%M:%S",
                           tz = "EST"),
          #scale temp to fit on same graph
          temp = temp *  0.0625 + 6.45) %>%
  group_by(hms = round_date(hms, "10 minutes"), tank) %>%
  summarise(pH.mean = mean(pH,na.rm=T),
            pH.sd = sd(pH,na.rm=T),
            temp.mean = mean(temp,na.rm=T),
            temp.sd = sd(temp,na.rm=T)) %>%
  ggplot(aes(hms)) +
  geom_smooth(aes(y=pH.mean, color="pH"),
              se = F, span=0.5) +
  geom_ribbon(aes(ymin=pH.mean-pH.sd,
                  ymax=pH.mean+pH.sd,
                  fill="pH"), alpha=0.3) + 
  geom_line(aes(y=temp.mean, color="Temp")) +
  geom_ribbon(aes(ymin=temp.mean-temp.sd,
                ymax=temp.mean+temp.sd,
                fill="temp"),alpha=0.3) + 
  scale_y_continuous(
    # Features of the first axis
    name = "pH",
    breaks = seq(7.8,8.2,by=0.2),
    minor_breaks = seq(7.7,8.2,by=0.05),
    limits = c(7.7,8.2),
    # Add a second axis and specify its features
    sec.axis = sec_axis(~(.- 6.45) / 0.0625, 
                        name="Temperature")
  ) + 
  scale_x_datetime("Hour",
                   limits = as.POSIXct(strptime(c("00:00",
                                                  "23:59"),
                                                format = "%H:%M", tz = "EST")),
                   breaks = seq(as.POSIXct("00:00", format="%H:%M",tz = "EST"),
                                as.POSIXct("24:00", format="%H:%M", tz = "EST"),
                                by = "3 hours"),
                   labels = date_format("%H"),
                   expand = c(0,0)) +
  #night/day boxes
  annotate("rect", xmin = as.POSIXct(strptime("19:00", format = "%H:%M", tz = "EST")),
                  xmax = as.POSIXct(strptime("23:59", format = "%H:%M", tz = "EST")),
                  ymin = -Inf, ymax = Inf, alpha = .2) + 
  annotate("rect", xmin = as.POSIXct(strptime("0:00", format = "%H:%M",tz = "EST")),
                xmax = as.POSIXct(strptime("7:00", format = "%H:%M", tz = "EST")),
                ymin = -Inf, ymax = Inf, alpha = .2) + 
  facet_wrap(~tank) +
  theme_bw() +
  guides(fill = FALSE)
```

Variability is still present, but the extreme spikes caused by cleaning have been removed. Thus, any variability that remains is due to durafet error or experimental variability that affected the corals. For example, the durafet for T15 had much higher variability than the other aquariums. However, I believe this to be negligible as values are not spiking out of the treatment groups.

## Carbonate Chemistry Data

500mL water samples were collected every Tuesday for analysis of the complete carbonate chemistry suite. 

```{r load data from google sheets}
ERL <- read_csv("~/Grad School/ULink 2022/carbonateChemistry/ERLcollectionData.csv")
```

```{r load data from carb database}
#load data from Bottle Database - Final_Query_with_Spec_pH
#one day will be SQL from Mike's DB
carbSystem <- read_csv("~/Grad School/ULink 2022/carbonateChemistry/bottleDB_export.csv") %>%
  filter(ID_Sample %in% ERL$tag) %>%
  select(tag = ID_Sample,
         DIC.kg = Avg_DIC_Kg_CORRECTED,
         TA.kg = Avg_TA_Kg_CORRECTED,
         sal = Calcdsalinity,
         pH.spec = CORRECTED_pH,
         spec.temp = Spec_pH_TEMP_C) %>%
  mutate_all(funs(as.numeric(.))) %>%
  left_join(ERL,
            by = "tag") %>%
  #temperature correct pH spec w/ ERL temp data
  mutate(pH.spec = pHinsi(pH = pH.spec,
                                    Tinsi = temp.ERL,
                                    Tlab = spec.temp)) %>%
  select(-c(spec.temp)) %>%
#solve carbonate system
  mutate(carb(flag = 15, #flag 15 for Alkalinity and DIC, respectively
       var1 = TA.kg/1000000, 
       var2 = DIC.kg/1000000,
       S = sal,
       T = temp.ERL,
       #use all default constants as recommended by Dickson et al. (2007) 
       k1k2 = "l",
       kf = "pf",
       ks = "d", 
       pHscale="T",
       b = "u74")) %>%
  select(tank,date,tag,
         temp.ERL,
         DIC.kg,TA.kg,
         pH.spec,pH.ERL,
         pH.solved=pH,pCO2,omega=OmegaAragonite,HCO3,CO3)
```

### Time of Day
The bottles were not taken at exactly the same time of day (my fault). Therefore there will be enhanced variability of these stats since our diel variability is in action.

```{r todSampling}
capFig("Times of Aquarium Bottle Sampling")
carbSystem %>%
  mutate(hms = as.POSIXct(format(date,
                           format="%H:%M:%S"), 
                           format = "%H:%M:%S",
                           tz = "America/New_York")) %>%
  group_by(date = date(date(date))) %>%
  summarise(hms = mean(hms,na.rm = T)) %>%
  ggplot(aes(date, hms)) +
  geom_point() +
  theme_bw() +
  labs(x = "Date",
       y = "Time of Day")
```

Sampling time had a mean of  around 10a. 3 sampling times were taken after 12p with one sampling time around 4:20p

## Carb Parameters
```{r carbsystem}
carbSystem %>%
  #only use data once treatment conditions met 4/1
  filter(date(date) > as.Date("2022-04-01")) %>%
  select(tank, temp = temp.ERL, TA = TA.kg, DIC = DIC.kg, pCO2, omega) %>%
  group_by(tank) %>%
  summarise(across(everything(),
                    ~ paste(format(round(mean(., na.rm=T),2),nsmall = 2),
        format(round(sd(.,na.rm = T)/sqrt(n()),2),nsmall = 2),
        sep = " ± "))) %>%
  kbl(align = 'c',
    digits = 4,
    caption = "Table 1: Aquarium Conditions. Means ± SEM") %>%
  kable_classic() %>%
  row_spec(0, bold = T)
  
```

## Statistics

```{r carb statistics}
carbTbl <- carbSystem %>%
  #only use data once treatment conditions met 4/1
  filter(date(date) > as.Date("2022-04-01")) %>%
  select(tank, temp = temp.ERL, TA = TA.kg, DIC = DIC.kg, pCO2, omega) %>%
  mutate(treatment = case_when(tank %in% c('T13','T15') ~ 'HCO2',
                               TRUE ~ 'LCO2')) %>%
  pivot_longer(temp:omega,
               names_to = "parameter") %>%
  group_by(parameter) %>%
  nest() %>%
  mutate(mod = map(data, ~aov(data = .,
                    formula = value ~ treatment/tank))) %>%
  mutate(model_tidy = map(mod, broom::tidy)) %>%
  unnest(model_tidy) %>%
  select(-c(data,mod)) %>%
  mutate(significance = case_when(p.value <0.001 ~ "xxx",
                                  p.value <0.01 ~ "xx",
                                  p.value <0.05 ~ "x",
                                  TRUE ~ "NS"))

carbTbl %>%
    kbl(align = 'c',
    digits = 3,
    caption = "Table 2: Significance of Parameters") %>%
  kable_classic() %>%
  row_spec(which(carbTbl$p.value<0.05),
           bold = T,
           color = "white",
           background = "red")
```

Temperature and total alkalinity were not significantly different between treatments or within treatments (p>>0.05). DIC, pCO2, and $$\Omega_{Ar}$$ (p<0.001) were significantly different between treatment, but not between aquariums within treatment (p>>0.05). In other words, our system reproducibly altered the carbonate chemistry parameters with high precision.


# Calcification Analysis

```{r load buoyant weight data}
#selet the authorized tokens
googledrive::drive_auth("p.kiel98@gmail.com")
gs4_auth("p.kiel98@gmail.com")

#load data from google drive
bw <- googledrive::drive_ls(path = "~/GradSchool/ULink_OAxGenotype/",
                            pattern = "buoyantWeight$") %>%
  read_sheet(col_types = "Dcdddc") %>%
  mutate(tag = as.character(tag)) %>%
  drop_na(mass)

#load tank assignment + genotype data
assignment <- read.csv('~/Grad School/ULink 2022/tankAssignment.csv') %>%
  filter(treatment != "") %>%
  mutate(tag = as.character(tag),
         tank = as.factor(tank))

bw <- bw %>%
  left_join(assignment, by="tag") %>%
  arrange(tank, genotype,tag,date) %>%
  filter(date > as.Date("2022-03-15")) %>%
#calculate weight in air
  mutate(
    #calculate sw density (g/cm3) from YSI, p taken as std atm p
    density = rho(S = salinity,
                  T = temp)/1000,
    #calculate weight in air using Jokiel et al. 1978 eqn
    mass_a = mass/(1-density/2.93)) %>%
  select(-c(mass,temp,salinity, density))

#corals to check
#weird growth
#10, 63, 103,
#likely broke
#120, 49

#calculate total growth
totalGrowth <- bw %>%
  #set starting date
  filter(date > as.Date("2022-04-15") & genotype != "Cheetos-B") %>%
  mutate(tag = as.numeric(tag)) %>%
    #add SA
  left_join(read.csv("C:/Users/pkiel/Documents/Grad School/ULink 2022/3dScans/initial/saVol_KielMarch22.csv") %>%
            select(tag, SA = `SA.cm2_corrected`,
                   V = `V.mm3_corrected`, SA_V, 
                   H = `H.cm_corrected`) %>%
           mutate(tag = as.numeric(str_extract(tag, "[[:digit:]]+"))),
          by="tag") %>%
  #remove standards
  drop_na(genotype,mass_a) %>%
  group_by(tag,genotype,tank, treatment, SA) %>%
  summarise(days = as.numeric(last(date)-first(date)),
            newGrowth.mg = (last(mass_a) - first(mass_a))*1000,
            initMass = first(mass_a),
            .groups="drop") %>%
  #mg cm^-2 day^-1
  mutate(G = newGrowth.mg/SA/days)
```


```{r calcification graphs}
#capFig("Avg Daily Growth by Treatment")
oa_G <- totalGrowth %>%
  ggplot(aes(treatment, G, fill=treatment)) +
  geom_boxplot() +
  labs(x = "Treatment",
       y = expression(paste("Avg Daily Growth Rate (mg  ", cm^-2, " ", day^-1, ")"))) +
  theme_classic()

#capFig("Avg Daily Growth by Genotype")
geno_G <-totalGrowth %>%
  filter(tag!=90) %>%
  ggplot(aes(genotype, G, fill=treatment)) +
  geom_boxplot() +
  labs(x = "Genotype") +
  theme_classic()

capFig("Avg Daily Growth by (A) Treatment and (B) Genotype")
ggarrange(oa_G, geno_G + rremove("ylab") + rremove("y.text"),
          labels = c("A", "B"),
          hjust = c(-0.5,-1.5),
          ncol = 2,
          widths = c(2,3),
          common.legend = T,
          legend = "right")
```

```{r calcification stats}
groupMeans <- totalGrowth %>%
  group_by(treatment) %>%
  get_summary_stats(G, type = "mean_sd")

groupMeans %>%
  select(-variable) %>%
      kbl(align = 'c',
    digits = 4,
    caption = "Table 2: Means of Treatment") %>%
  kable_classic() %>%
  row_spec(0, bold = T)

tTestResults <- totalGrowth %>% 
  t_test(G ~ treatment, var.equal = T) %>%
  add_significance()

tTestResults %>%
  kbl(align = 'c',
    digits = 4,
    caption = "Table 3: Stats of Treatment",
    escape=F) %>%
  kable_classic() %>%
  row_spec(0, bold = T)

cohensResults <- totalGrowth %>% 
  cohens_d(G ~ treatment, var.equal = T) 

cohensResults %>%
      kbl(align = 'c',
    digits = 4,
    caption = "Table 4: Effect Size of Treatment") %>%
  kable_classic() %>%
  row_spec(0, bold = T)

#fixed effects model
mod <- totalGrowth %>%
  aov(G ~ treatment*genotype,
                 data=.)

tukeyGTbl <- mod %>%
  TukeyHSD() %>%
  broom::tidy() %>%
    filter(contrast %in% c('LCO2-HCO2',
                        'MB-C-AC-2,',
                        'SI-A-AC-2',
                        'SI-A-MB-C',
                        'LCO2:AC-2-HCO2:AC-2',
                        'LCO2:MB-C-HCO2:MB-C',
                        'LCO2:SI-A-HCO2:SI-A')) %>%
  select(term,contrast,adj.p.value) %>%
  mutate(significance = case_when(adj.p.value <0.001 ~ "xxx",
                                  adj.p.value <0.01 ~ "xx",
                                  adj.p.value <0.05 ~ "x",
                                  TRUE ~ "NS"),
         adj.p.value = round(adj.p.value, digits=4)) %>%
  arrange(adj.p.value)

tukeyGTbl %>%
  kbl(align = 'c',
    digits = 4,
    caption = "Table 5: TukeyHSD Results of Anova",
    escape = TRUE) %>%
  kable_classic() %>%
  row_spec(0, bold = T) %>%
  row_spec(which(tukeyGTbl$adj.p.value<0.05),
         bold = T,
         color = "white",
         background = "red")
  
# multcomp::cld(emmeans(mod,
#                       list(pairwise ~ treatment*genotype),
#                       adjust = "tukey"),
#               Letters = letters) %>%
#   select(treatment, genotype, significance = .group) %>%
#     kbl(align = 'c',
#     digits = 4,
#     caption = "Table 6: Significance Letter Groups") %>%
#   kable_classic() %>%
#   row_spec(0, bold = T)
```

The mean calcification rate in the HCO2 group was mean `r round(mean(totalGrowth$G[which(totalGrowth$treatment=='HCO2')]),3)`  (SD = `r round(sd(totalGrowth$G[which(totalGrowth$treatment=='HCO2')]),3)`) mg/ $$cm^2$$ /day, whereas the mean in the LCO2 group was `r round(mean(totalGrowth$G[which(totalGrowth$treatment=='LCO2')]),3)` (SD = `r round(sd(totalGrowth$G[which(totalGrowth$treatment=='LCO2')]),3)`). A Student two-samples t-test showed that the difference was statistically significant, t(`r tTestResults$df`) = `r round(tTestResults$statistic,3)`, p < 0.0001, d = `r round(cohensResults$effsize,3)`. Thus, the ocean acidification group saw on average a `r round((groupMeans$mean[2]-groupMeans$mean[1])/groupMeans$mean[2]*100,0)`% reduction in calcification rates. The effects, however, were not even across the genotypes (Table 5).

## Tank Effects

We saw above that tank conditions were significantly different among treatment groups, but not individual aquariums within treatment. We also saw that calcification rates were significantly different among treatment groups. Here I am analyzing tank effects on the calcification rate and investigating if calcification rates were significantly different between aquariums within the same treatment group. This will dictate whether we need to include tank as a random intercept in our ANOVA models.

```{r tank effects graph}
manLegend <- c("13 HCO2"= "#f0786f",
              "14 LCO2"= "#3dbec3",
              "15 HCO2"= "#f0786f",
              "16 LCO2" = "#3dbec3")

capFig("Avg Daily Growth by Tank and Treatment")
totalGrowth %>%
  ggplot(aes(tank, G, fill=treatment)) +
  geom_boxplot() +
  labs(x = "Genotype",
       y = expression(paste("Avg Daily Growth Rate (mg  ", cm^-2, " ", day^-1, ")"))) +
  theme_classic()
# 
# capFig("Avg Daily Growth by Genotype, Treatment, and Tank")
# totalGrowth %>%
#   mutate(tank = case_when(tank == 13 ~ "13 HCO2",
#                           tank == 14 ~ "14 LCO2",
#                           tank == 15 ~ "15 HCO2",
#                           tank == 16 ~ "16 LCO2")) %>%
#   ggplot(aes(genotype, G)) +
#   geom_boxplot(aes(fill=tank)) +
#   scale_fill_manual(values = manLegend) +
#   labs(x = "Genotype",
#        y = expression(paste("Avg Daily Growth Rate (mg  ", cm^-2, " ", day^-1, ")"))) + 
#   theme_classic()
```

### Tank Effects Statistics

```{r tank effects statistics}
#t-test between tanks
totalGrowth %>%
  group_by(treatment) %>%
  t_test(data=.,formula = G ~ tank) %>%
    kbl(align = 'c',
    digits = 3,
    caption = "Table 6: Significance testing of treatment effect on calcification, using t test") %>%
  kable_classic() %>%
  row_spec(0, bold = T)

genos <- unique(totalGrowth$genotype)
pairwiseString <- c(paste0('(',
                         genos,
                         ' HCO2) - (',
                         genos,
                         ' LCO2)'),
                    paste0('(HCO2 ',
                         genos,
                         ') - (LCO2 ',
                         genos,
                         ')'))
```

No observable differences of mean calcification rate when comparing within treatment groups. 

### Mixed Effects Model

Here I created a mixed effects model model to account for the lack of independence brought upon by having multiple corals grown in the same tank and the possible tank-specific effects that may have affected calcification rates. Including this random effect increased the AIC from -354 to -352 as compared to a fixed-effects only model, and therefore I will not include random tank effects in my analysis.

As a reminder, here is the fixed effects model as shown above:

```{r fixed effects reminder, echo=T}
modFixed <- totalGrowth %>%
  aov(G ~ genotype*treatment, data=.)

modFixed %>%
  summary()
```

Here is the mixed effects model with the tank identity set as a random factor to give each tank its own, random intercept. Notice, including the random effects decreases the absolute value of the AIC. Therefore, this new model better describes the data.
```{r random mixed effects, echo=T}
modRandom <- totalGrowth %>%
  lmerTest::lmer(G ~ genotype * treatment + (1|tank),
                 data=.)
modRandom %>%
  anova() %>%
  tidy() %>%
  kbl()

#redefining modRandom w/ REML=F for AIC comparison
modRandom <- totalGrowth %>%
  lmerTest::lmer(G ~ genotype * treatment + (1|tank), REML = F,
                 data=.)

AIC(modFixed, modRandom) %>%
  kbl()
```

<!-- Tukey post-hoc analysis of the mixed effects model: -->

<!-- ```{r random mixed effects table} -->
<!-- emmeans(modRandom, list(pairwise ~genotype*treatment), -->
<!--         adjust = "tukey")[[2]] %>% -->
<!--   broom::tidy() %>% -->
<!--   filter(`1` %in% pairwiseString) %>% -->
<!--   select(-c(term,df)) %>% -->
<!--   rename(pairwise=`1`) %>% -->
<!--   kbl(align = 'c', -->
<!--     digits = 4, -->
<!--     caption = "Table 8: Pairwise comparison of genotypes' sensitivity to OA", -->
<!--     escape=F) %>% -->
<!--   kable_classic() %>% -->
<!--   row_spec(0, bold = T) %>% -->
<!--   row_spec(c(1,3), -->
<!--          bold = T, -->
<!--          color = "white", -->
<!--          background = "red") -->
<!-- ``` -->

<!-- Pattern is the same as above with the fixed effects. Significance has decreased (p values increased) for AC-2 and MB-C, suggesting that there was some variability between tanks in the same treatment (T13 v T15 and T14 v T16) but that this within treatment variability was not significant enough to change our conclusions. Thus, we fail to reject our null hypothesis that there are significant differences between individual genotype's susceptibility to OA. -->

## Powder Available

```{r powder available}
capFig("Amount of New CaCO3 Precipitated")
totalGrowth %>%
  ggplot(aes(x=genotype, y=newGrowth.mg)) +
  geom_point(aes(color=treatment)) +
  scale_y_continuous(limits = c(0,1000)) +
  geom_hline(yintercept = 50, color="red") +
  geom_hline(yintercept = 120, color="orange") +
  geom_hline(yintercept = 500, color="green") +
  labs(y = "Precipated Skeleton (mg)") +
  theme_classic()
```

The amount of new aragonite precipated is visualized above. Horizontal lines denote thresholds for tests: >500mg = green (complete suite including XRD), >120 mg = orange (complete suite sans XRD), >50mg = red (TGA and isotope only).

# Linear Extension

```{r import LE data}
height <- read.csv('~/Grad School/ULink 2022/initialHeights.csv') %>%
  mutate(date=as.Date("2022-02-25")) %>%
  select(tag,height=height.mm,date) %>%
  left_join(read.csv('~/Grad School/ULink 2022/height06102022.csv') %>%
              mutate(date=as.Date("2022-06-10"),
                     .after="height"),
            by="tag") %>%
  select(tag:date.y) %>%
  left_join(read.csv('~/Grad School/ULink 2022/height07152022.csv'),
            by="tag") %>%
  mutate(height.z=height,
         date.z=as.Date("2022-07-15")) %>%
  select(-height) %>%
  pivot_longer(!tag,
               names_to = '.value',
               names_pattern = '(.*)\\.',
               values_drop_na = T) %>%
  drop_na() %>%
  arrange(tag,date) %>%
  group_by(tag) %>%
  summarise(init.mm = min(height),
            fin.mm = max(height),
            days = as.numeric(date[which.max(height)]-first(date))) %>%
  filter(days!=0) %>%
  mutate(prod = (fin.mm-init.mm)/(init.mm/10)/days,
         LE = fin.mm-init.mm) %>%
  left_join(assignment %>%
              mutate(tag = as.numeric(tag)),
            by="tag") %>%
  filter(genotype!="Cheetos-B")

```

```{r LE graphs}
capFig("Avg Daily Linear Extension by Treatment")
height %>%
  ggplot(aes(treatment, prod, fill=treatment)) +
  geom_boxplot() +
  labs(x = "Treatment",
       y = "Avg Daily LE Rate (mm/cm/day)") +
  theme_classic()

capFig("Avg Daily Linear Extension by Treatment and Aquarium")
height %>%
  ggplot(aes(tank, prod, fill=treatment)) +
  geom_boxplot() +
  labs(x = "Treatment",
       y = "Avg Daily LE Rate (mm/cm/day)") +
  theme_classic()

capFig("Avg Daily Linar Extension by Genotype and Treatment")
height %>%
  ggplot(aes(genotype, prod)) +
  geom_boxplot(aes(fill=treatment)) +
  labs(x = "Genotype",
       y = "Avg Daily LE Rate (mm/cm/day)") +
  theme_classic()
```

Such large variability in that SI-A ambient group.

```{r LE stats}

groupMeans <- height %>%
  group_by(treatment) %>%
  get_summary_stats(prod, type = "mean_sd")

groupMeans %>%
  select(-variable) %>%
      kbl(align = 'c',
    digits = 4,
    caption = "Table 7:Means of Treatment") %>%
  kable_classic() %>%
  row_spec(0, bold = T)

tTestResults <- height %>% 
  t_test(prod ~ treatment, var.equal = T) %>%
  add_significance()

tTestResults %>%
  kbl(align = 'c',
    digits = 4,
    caption = "Table 8: Stats of Treatment",
    escape=F) %>%
  kable_classic() %>%
  row_spec(0, bold = T)

cohensResults <- height %>% 
  cohens_d(prod ~ treatment, var.equal = T) 

cohensResults %>%
      kbl(align = 'c',
    digits = 4,
    caption = "Table 9: Effect Size of Treatment") %>%
  kable_classic() %>%
  row_spec(0, bold = T)

#fixed effects model
mod <- height %>%
  aov(prod ~ treatment*genotype,
                 data=.) 

tukeyLE <- mod %>%
  TukeyHSD() %>%
  broom::tidy() %>%
  filter(contrast %in% c('LCO2-HCO2',
                        'MB-C-AC-2,',
                        'SI-A-AC-2',
                        'SI-A-MB-C',
                        'LCO2:AC-2-HCO2:AC-2',
                        'LCO2:MB-C-HCO2:MB-C',
                        'LCO2:SI-A-HCO2:SI-A')) %>%
  select(term,contrast,adj.p.value) %>%
  mutate(significance = case_when(adj.p.value <0.001 ~ "xxx",
                                  adj.p.value <0.01 ~ "xx",
                                  adj.p.value <0.05 ~ "x",
                                  TRUE ~ "NS"),
         adj.p.value = round(adj.p.value, digits=4)) %>%
  arrange(adj.p.value)

tukeyLE %>%
    kbl(align = 'c',
    digits = 4,
    caption = "Table 10: TukeyHSD Results of Anova",
    escape = F) %>%
  kable_classic() %>%
  row_spec(0, bold = T) %>%
  row_spec(which(tukeyLE$adj.p.value<0.05),
       bold = T,
       color = "white",
       background = "red")

# multcomp::cld(emmeans(mod,
#                       list(pairwise ~ treatment*genotype),
#                       adjust = "tukey"),
#               Letters = letters) %>%
#   select(treatment, genotype, significance = .group) %>%
#     kbl(align = 'c',
#     digits = 4,
#     caption = "Table 13: Significance Letter Groups") %>%
#   kable_classic() %>%
#   row_spec(0, bold = T)
```

The mean linear extension rate in the HCO2 group was mean `r round(mean(height$prod[which(height$treatment=='HCO2')]),3)`  (SD = `r round(sd(height$prod[which(height$treatment=='HCO2')]),3)`) mm/cm/day, whereas the mean in the LCO2 group was `r round(mean(height$prod[which(height$treatment=='LCO2')]),3)` (SD = `r round(sd(height$prod[which(height$treatment=='LCO2')]),3)`). A Student two-samples t-test showed that the difference was not statistically significant, t(`r tTestResults$df`) = `r round(tTestResults$statistic,3)`, p =0.124, d = `r round(cohensResults$effsize,3)`. 

<!-- ## Tank Effects -->

<!-- ```{r LE tank effects} -->
<!-- capFig("Avg Daily Linear Extension by Treatment and Aquarium") -->
<!-- height %>% -->
<!--    mutate(tank = case_when(tank == 13 ~ "13 HCO2", -->
<!--                           tank == 14 ~ "14 LCO2", -->
<!--                           tank == 15 ~ "15 HCO2", -->
<!--                           tank == 16 ~ "16 LCO2")) %>% -->
<!--   ggplot(aes(genotype, prod, fill=tank)) + -->
<!--   geom_boxplot() + -->
<!--   scale_fill_manual(values = manLegend) + -->
<!--   labs(x = "Treatment", -->
<!--        y = "Avg Daily LE Rate (mm/cm/day)") + -->
<!--   theme_classic() -->

<!-- ``` -->

<!-- ```{r LE Tank effects stats} -->
<!-- height %>% -->
<!--   group_by(treatment) %>% -->
<!--   t_test(data=.,formula = prod ~ tank) %>% -->
<!--     kbl(align = 'c', -->
<!--     digits = 3, -->
<!--     caption = "Table 14: Significance testing of tank effect on Linear Extension Rates") %>% -->
<!--   kable_classic() %>% -->
<!--   row_spec(0, bold = T) -->

<!-- modFixed <- height %>% -->
<!--   aov(prod ~ treatment*genotype, -->
<!--                  data=.)  -->

<!-- modFixed %>% TukeyHSD() %>% -->
<!--   broom::tidy() %>% -->
<!--   filter(contrast %in% c('LCO2:SI-A-HCO2:SI-A')) %>% -->
<!--   select(term,contrast,adj.p.value) %>% -->
<!--   mutate(significance = case_when(adj.p.value <0.001 ~ "xxx", -->
<!--                                   adj.p.value <0.01 ~ "xx", -->
<!--                                   adj.p.value <0.05 ~ "x", -->
<!--                                   TRUE ~ "NS"), -->
<!--          adj.p.value = round(adj.p.value, digits=4)) %>% -->
<!--   arrange(adj.p.value) -->


<!-- modRandom <- height %>% -->
<!--   lmerTest::lmer(prod ~ genotype * treatment + (1|tank), -->
<!--                  data=.) -->
<!-- modRandom %>% -->
<!--   anova() -->

<!-- emmeans(modRandom, list(pairwise ~genotype*treatment), -->
<!--         adjust = "tukey")[[2]] %>% -->
<!--   broom::tidy() %>% -->
<!--   filter(`1` %in% pairwiseString) %>% -->
<!--   select(-c(term,df)) %>% -->
<!--   rename(pairwise=`1`) %>% -->
<!--   kbl(align = 'c', -->
<!--     digits = 4, -->
<!--     caption = "Table 15:Pairwise comparison of genotypes' sensitivity to OA", -->
<!--     escape=F) %>% -->
<!--   kable_classic() %>% -->
<!--   row_spec(0, bold = T) %>% -->
<!--   row_spec(4, -->
<!--          bold = T, -->
<!--          color = "white", -->
<!--          background = "red") -->

<!-- AIC(modFixed, modRandom) -->


<!-- ``` -->

<!-- AIC tells us the mixed effects model better describes the data. Post-hoc testing further tells us that indeed SI-A's linear extension rates were significantly different between treatments (p<0.05), yet all other genotype's were not. -->

<!-- ```{r total extension} -->
<!-- capFig("Amount of New Skeleton and Mean New LE (Red Line)") -->
<!-- height %>% -->
<!--   ggplot(aes(x=genotype, y=LE)) + -->
<!--   geom_point(aes(color=treatment)) + -->
<!--   scale_y_continuous(limits = c(0,30)) + -->
<!--   geom_hline(yintercept = mean(height$LE), -->
<!--              color="red") + -->
<!--   labs(y = "Linear Extension (mm)") + -->
<!--   theme_classic() -->
<!-- ``` -->

# Takeaways and Next Steps

Overall growth was less than hoped for. However, there is enough new skeleton for nearly all the powder tests that we want to conduct: FTIR, TGA, boron isotopes, and Raman.For nanoindentation tests, we should have enough new growth to run on a majority of samples.

Linear extension was measured by maximum vertical height as measured with calipers. We additionally have initial 3d scans of all corals and post 3d scans of a subset of 48 corals (n=3 per genotype per tank). From this data, we can extract surface area to volume ratios and see how this changed among genotypes and treatments. This analysis still needs to be done. 

## CT Scanning
The micro-ct scan is currently out of service. We can use the large ct-scanner to determine bulk density. The scanner has a resolution of 0.1mm/scan so we can measure the density of the new growth. This growth is isolated to the highly variable apical tips which may cause some problems. <a href="https://patrickmkiel.com/notebook/research/LangdonSkeletonAnalysis/?pass=acidification" target="_blank">See this post which discusses the ct-scanning analysis of apical tips done on Langdon's OA corals.</a> 


</div>
